---
title: "Horizontal Integration"
author: "Nick Isaac"
date: "2024-07-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(lme4)
library(nimble)
library(dplyr)
library(ggmcmc)
library(coda)
```



Read in the Expert Elicitation
```{r}
experts <- readRDS("data/Hymenoptera_experts_fertiliser_pesticide.rds")
str(experts)
```

PLot the Expert data
```{r}
ggplot(data = experts, aes(x=Threat_intensity, y=1+Best_estimate)) +
  geom_line(aes(col=factor(Expert))) +
  facet_wrap(~Threat_response, scales = "free") +
  geom_ribbon(aes(ymin=1+Lowest, ymax=1+Highest, fill = factor(Expert)), alpha=0.1) +
  scale_y_log10() +
  ylab("Abundance") +
  theme_bw() +
  ggtitle("Expert Elicitation")
```

Simple model of EE data
```{r}
emod <- lmer(log(1+Best_estimate) ~ poly(Threat_intensity,2) + (1|Expert), 
             data = subset(experts, 
                           Threat_response == "Agri & forest effluent PEST Ab"))
#summary(emod)

# replace with explicit poly terms
emod2 <- lmer(log(1+Best_estimate) ~ Threat_intensity + I(Threat_intensity^2) + (1|Expert), 
             data = subset(experts, 
                           Threat_response == "Agri & forest effluent PEST Ab"))
#summary(emod2)


# simple linear model
emod3 <- lmer(log(1+Best_estimate) ~ Threat_intensity + (1|Expert), 
             data = subset(experts, 
                           Threat_response == "Agri & forest effluent PEST Ab"))
summary(emod3)
```

Explore fitted values
```{r}

```



Read in the PREDICTS model estimates
```{r}
#predicts <- readRDS("data/Hymenoptera_PREDICTS_pesticide.rds")
predicts <- readRDS("data/PREDICTS_plus_threats_byLU_scaled.rds")
#str(predicts)
```

Plot the PREDICTS data
```{r}
ggplot(data = predicts, aes(x=pest_avg_kg_sqrt_RS, y=logAbundance)) +
  geom_line(aes(col=SSB)) +
  facet_wrap(~Predominant_habitat, scales = "free") +
  theme_bw() +
  ylab("Abundance") +
  theme(legend.position = "none") +
  ggtitle("PREDICTS")
```
Fit the PREDICTS model
```{r}
pmod <- lmer(logAbundance ~ poly(pest_avg_kg_sqrt_RS,2) + Predominant_habitat +
               (1|SS) + (1|SSB), 
             data = subset(predicts, !is.na(pest_avg_kg_sqrt_RS)))
```

# Edit
```{r manipulate data}
predicts2 <- predicts %>%
  filter(!is.na(pest_avg_kg_sqrt_RS)) %>% 
  dplyr::mutate(pest = (pest_avg_kg_sqrt_RS * 0.315872471538644 + 0.217945983339326)^2) %>% 
  dplyr::mutate(pest = ifelse(pest < 0.00001, 0, pest)) %>%
  dplyr::mutate(SS = as.numeric(SS)) %>%
  dplyr::mutate(SSB = as.numeric(SSB)) %>%
  select(SS, SSB,
         Predominant_habitat,
    logAbundance, 
    pest)

# need to convert Predominant_habitat into dummy variables

experts2 <- experts %>%
  filter(Threat_response == "Agri & forest effluent PEST Ab") %>%
  #dplyr::mutate(logAbundance = log(1+Best_estimate)) %>%
  #dplyr::filter(Expert == 2) %>% # temp to simplify data
  dplyr::select(Threat_intensity,
                Expert, Best_estimate)
```


```{r replot}
ggplot(data = predicts2, aes(x=pest, y=logAbundance)) +
  geom_line(aes(col=SSB)) +
  facet_wrap(~Predominant_habitat, scales = "free") +
  theme_bw() +
  theme(legend.position = "none") +
  ggtitle("PREDICTS")

ggplot(data = experts2, aes(x=Threat_intensity, y=1+Best_estimate)) +
  geom_line(aes(col=factor(Expert))) +
#  geom_ribbon(aes(ymin=1+Lowest, ymax=1+Highest, fill = factor(Expert)), alpha=0.1) +
  theme_bw() +
  scale_y_log10() +
  ylab("Abundance") +
  ggtitle("Expert Elicitation")
```





# Model
Try a JL model. We have to model the two datasets in the same way. This is not ideal. PREDICTS data is typically transformed before use. 

We'll subtract 1 from the threat scores to help convergence and identifiability.
```{r defineModel}
modelcode <- nimbleCode({ 
  #### we fit a model with shared parameters 
 if(runPredicts){ 
  #### Predicts sub-model
  for(i in 1:nP){
    exp(logN_P[i]) ~ dpois(e_N_P[i])
    log(e_N_P[i]) <- alpha + beta1 * (threat_P[i]-1) + 
      beta2 * threat_P[i]^2 +
      eta1[SS[i]] + eta2[SSB[i]]
       #gamma1 * Hab[i] # categories - need to code dummies
  }
 }
  #### EE sub-model
  for(j in 1:nE){
    N_E[j] ~ dpois(e_N_E[j])
    log(e_N_E[j]) <- alpha + beta1 * (threat_E[j]-1) + 
      #beta2 * threat_E[j]^2 + 
      eta3[Expert[j]]
  }
  
  # random effects
if(runPredicts){ 
  for (k in 1:nSS) {eta1[k] ~ dnorm(0, sd = sigma1)}
  for (k in 1:nSSB) {eta2[k] ~ dnorm(0, sd = sigma2)}
}
  for (k in 1:nExpert) {eta3[k] ~ dnorm(0, sd = sigma3)}
  
  # priors
  #alpha ~ T(dt(0, 1, 1), 0, Inf) # Half Cauchy 
  #alpha ~ dgamma(1, 1)
  alpha ~ dnorm(3, tau = 1/2.71) 
  beta1 ~ dnorm(0, tau = 1/2.71)
  beta2 ~ dnorm(0, tau = 1/2.71)
  
  # hyperprior
if(runPredicts){ 
  sigma1 ~  T(dt(0, 1, 1), 0, Inf) # Half Cauchy
  sigma2 ~  T(dt(0, 1, 1), 0, Inf) # Half Cauchy
  }
  sigma3 ~  T(dt(0, 1, 1), 0, Inf) # Half Cauchy

  })
```



Now fit the model for the Expert Elicitation only
```{r runModel}
runPredicts <- FALSE

# step 1 create an operational from from NIMBLE/JAGS/BUGS code
model <- nimbleModel(code = modelcode, 
                     constants = list(nE = nrow(experts2),
                                      nExpert = max(experts2$Expert),
                                      threat_E = experts2$Threat_intensity,
                                      Expert = experts2$Expert
                     ),
                     data = list(N_E = experts2$Best_estimate), 
                     inits = list(alpha = runif(1, 2, 5),
                                  beta1 = rnorm(1, 0, 1),
                                  beta2 = rnorm(1, 0, 1),
                                  sigma3 = rpois(1, 5))
)
#model$getVarNames()
```


```{r}
#model$getNodeNames()
model$initializeInfo()
#model$plotGraph()
#model$logProb_alpha
```


```{r}
# step 2 build an MCMC object using buildMCMC(). we can add some customization here
occMCMC <- buildMCMC(model, 
                     monitors = c('alpha', 
                                  "beta1",# "beta2",
                                  #"sigma1", "sigma2", 
                                  "sigma3"
                                  ), 
                     thin = 3, 
                     useConjugacy = FALSE) # useConjugacy controls whether conjugate samplers are assigned when possible
#about 5 seconds

# step 3 before compiling the MCMC object we need to compile the model first
Cmodel <- compileNimble(model)
# 25 seconds (less for fewer nodes)

Cmodel$calculate() # if NA there is a problem

# now the MCMC (project = NIMBLE model already associated with a project)
CoccMCMC <- compileNimble(occMCMC, project = model)
```

Now run the model
```{r}
nit <- 3e6
# and now we can use either $run or runMCMC() on the compiled model object.
system.time(
  runMCMC_samples <- runMCMC(CoccMCMC, 
                             nburnin = round(nit*0.8), niter = nit, 
                             nchains = 3, thin = 10,
                             samplesAsCodaMCMC = T)
) 
```

## Examine the model output
Plot the MCMC traces
```{r diagnoseModel, fig.height = 10, fig.width=7}
plot1a <- runMCMC_samples %>% ggs() %>% ggs_traceplot(greek = TRUE)
plot1b <- runMCMC_samples %>% ggs() %>% ggs_density(greek = TRUE)
gridExtra::grid.arrange(plot1a, plot1b, ncol=2)
```

Simple summary statistics from IDM
```{r}
pars <- as.data.frame(summary(runMCMC_samples)$statistics)
#pars$name <- rownames(pars)
Rhat <- gelman.diag(runMCMC_samples)$psrf[,1]

print(cbind(pars[,1:2], Rhat))
```
`alpha` is not converged unless niter is very large (>1e7). Might be easier if we centered the covariate on zero.


from simple mixed effect model
```{r}
sd_RE <- as.numeric(VarCorr(emod3)[[1]])^(1/2)

temp <- rbind(summary(emod3)$coef[,1:2],
              c(sd_RE, NA))

rownames(temp) <- c("alpha", "beta", "sigma3")
print(as.data.frame(temp))
```
Parameter estimates are pretty close!

# model with PREDICTS
```{r runModel}
runPredicts <- TRUE

# step 1 create an operational from from NIMBLE/JAGS/BUGS code
model <- nimbleModel(code = modelcode, 
                     constants = list(nP = nrow(predicts2), 
                                      nE = nrow(experts2),
                                      nExpert = max(experts2$Expert),
                                      nSS = max(predicts2$SS), # need to renumber from 1:n
                                      nSSB = max(predicts2$SSB), # need to renumber from 1:n
                                      threat_P = predicts2$pest[,1], # rescale!
                                      threat_E = experts2$Threat_intensity,
                                      #Hab = predicts2$Predominant_habitat,
                                      Expert = experts2$Expert,
                                      SS = predicts2$SS,
                                      SSB = predicts2$SSB
                     ),
                     data = list(logN_P = predicts2$logAbundance, 
                                 N_E = experts2$Best_estimate), 
                     inits = list(alpha = runif(1, 2, 5),
                                  beta1 = rnorm(1, 0, 1),
                                  beta2 = rnorm(1, 0, 1),
                                  sigma1 = rpois(1, 5),
                                  sigma2 = rpois(1, 5),
                                  sigma3 = rpois(1, 5))
)
#model$getVarNames()
```

Now repeat other steps ....
